---
title: " ASVs with Phyloseq"
author: "Mar Schmidt"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---


# Goals 

The goal is to make a phylogenetic tree 

1. load in preprocessed phyloseq object. 
2. Create ASV fasta file from the phyloseq object
3. Align the 16S Sequences from our fasta file 
4. Create a maximum likelihood tree. 

## Inputs 
1. `raw_preprocessed_physeq`

## Outputs 
1. MAAFT_aligned_ASV.fasta
2. ASVs_unrooted.tree


# Setup 

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set Seed 
```{r set-seed}
set.seed(238438)
```

## Load Packages 
```{r load-packages}
# Install phytools
remotes::install_github("liamrevell/phytools")

# Install ggtree
devtools::install_github("GuangchuangYu/ggtree")

# Now load them into the session 
pacman::p_load(phytools, ggtree, tidyverse, install = FALSE)
```

## Load Data 

```{r load-physeq}
# Load physeq 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

# Write ASV Fasta 

```{r asv-fasta}
# Pull out ASV sequences and ASV names
asv_seq_df <- 
  raw_preprocessed_physeq@tax_table %>% 
  data.frame() %>% 
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

# Add the ">" to make fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

#View(asv_seq_df)

# Create fasta object
asv_seq_fasta <- c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseqs))
head(asv_seq_fasta)

# Write to a fasta file 
write(asv_seq_fasta, 
      file = "data/04_Phylogenetic_Tree/preprocessed_ASVs.fasta")

```


# Run Alignment with MAAFT 

Note that `engine.opts = '-l'` in our R chunk header name gives us our original terminal base environment. 


To test shell code via RMarkdown: 
- **Mac:** `command` + `option` + `enter` will send line-by-line code 
- **Windows:** `control` + `alt` + `enter`


```{r run-mafft, engine = 'bash', engine.opts = '-l'}
# Write bash code to run MAFFT
# First provide the path to MAFFT
export PATH=/programs/mafft/bin:$PATH

# change directories to provide the fasta file we made above 
cd data/04_Phylogenetic_Tree/
pwd

# Where am I? 
echo "The working directory is $PWD"

# Set a seed for consistency and reproducibility 
RANDOM=238428

# Now, actually run MAFFT
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta

# Where am I? 
echo "The working directory is $PWD"
```

# FastTree2

```{r run-fasttree2, engine = 'bash', engine.opts = '-l'}
# Where am I? 
echo "The working directory is $PWD"

# Load fasttree 
# Provide export path to fasttree 
export PATH=/programs/FastTree-2.1.11:$PATH

# change directories to provide the fasta file we made above 
cd data/04_Phylogenetic_Tree/
echo "The working directory is $PWD"

# Run Fasttree to generate phylogenetic tree 
# parameters: 
    # -nt = indicates it's a nucleotide alignment
    # -gtr = generalized time reversible substitution model 
    # -fastest speed up the model, reduce memory usage (recommended for datasets that have >50,000)
    # -log = output a log file 
    # input alignment file 
    # specify the output tree file 
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree

# Change back to the main directory 
cd ../../
echo "The working directory is $PWD"
```


# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```
