---
title: "Biodiversity across a salinity gradient"
author: "Marian Schmidt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../figures/05_Biodiversity")
```

# Goals 

Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the hill numbers as it relates to stations and salinity (in PSU).
4. Answer the scientific question at hand! 

## Inputs 

1. `phytree_preprocessed_physeq`

## Outputs 

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to visualize these changes in biodiversity
3. Statistical tests to actually show the changes and patterns in biodiversity.


# Set up the Environment 

## Set the seed 
```{r set-seed}
set.seed(238428)
```


## Load Packages & Colors 

```{r load-packages-colors}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               install = FALSE)

# Load in colors for plotting 
source("code/colors.R")
```

## Load in Data 

```{r load-data}
load("data/04_PhylogeneticTree/phytree_preprocessed_physeqs.RData")

# take a look at it! 
archaeal_rooted_physeq

# Intuition check on seq depth
min(sample_sums(archaeal_rooted_physeq))

# Make a metadata dataframe 
metadata_df <- 
  archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

# view of it
glimpse(metadata_df)
```

# Calculate Biodiversity with iNEXT 

```{r iNEXT-biodiversity}
# prepare the input data 
# Species in rows and the samples in columns 
iNEXT_input_df <- 
  archaeal_rooted_physeq %>%
  otu_table() %>%
  data.frame()

# Inpect 
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

# Run iNEXT 
# Remember to set the seed!
# ASVs in ROWs, samples in COLUMNS 
#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                    q = c(0, 1, 2),
#                    datatype = "abundance")
```

Since iNEXT takes a while to run, we will load in the data that was caluculated before! 

```{r load-iNEXT-data}
load("data/05_Biodiversity/iNEXT_data.RData")

# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)
```


# Plot the Rarefaction Curves 

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)


# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(station_colors = station_colors,
            station = names(station_colors)),
            by = "station")

head(iNEXT_color_df)
```


Remember the output of iNEXT

q = 0, richness
q = 1, shannon
q = 2, simpson

```{r plot-iNEXT-rarefaction}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$station_colors) + 
  scale_fill_manual(values = iNEXT_color_df$station_colors) + 
  scale_shape_manual(values = base::rep(17, nsamples(archaeal_rooted_physeq))) + 
  theme_bw() + 
  theme(legend.position = "none") + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") 
```

# Manually Plot Rarefaction 

```{r manual-rarefaction}
str(iNEXT_data)

iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # fix the samples to merge with metadata 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with the metadata
  left_join(., metadata_df, by = "names") %>%
  mutate(station = factor(station, levels = c("Copano West",
                                              "Copano East",
                                              "Mesquite Bay",
                                              "Aransas Bay",
                                              "Shipping Channel")))

# Inspection
dim(iNEXT_manual_df)

# Manually plot rarefaction curves 
iNEXT_manual_df %>%
  # Filter out rows that are calculated with actual sequences (no extrapolated)
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot! 
  ggplot(aes(x = m, y = qD, color = station, group = names)) + 
  geom_line() + 
  # Facet by station 
  facet_grid(Order.q~station, scales = "free") + 
  scale_color_manual(values = station_colors) + 
  theme_bw() + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") + 
  theme(legend.position = "bottom")
```

# Boxplot: Diversity by Station 

## Categorical Analysis 

```{r make-obs-df}
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") 

# check it
glimpse(obs_div_df)

# Plot boxplots by station against diversity
obs_div_df %>%
  ggplot(aes(x = station, y = qD, fill = station, 
             color = station)) + 
  facet_wrap(.~Order.q, scales = "free_y") + 
  geom_jitter() + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1, 
                                   vjust = 1))
```


# Continuous Relationships - Salinity (PSU) vs Biodiversity

## Environmental Variables 

```{r environ-check}
# Create dataframe of environmental variables 
env_df <- 
  metadata_df %>%
  dplyr::select(names, water_tempC:DO_mgL)

# Pair-wise interactions between env var
pairs(dplyr::select(env_df, -names), upper.panel = NULL)
```


## PSU 

- "Practical Salinity Unit"
- 1 PSU = 1 g salt per 1,000 g water = 1 ppt

ranges between 0 and >30

- Freshwater: 0-0.5
- Brackish: 0.5 - 30
- Marine: >30

```{r plot-div-salinity-lm}
# How does biodiversity change over the salinity gradient?
obs_div_df %>%
  ggplot(aes(x = salinity_psu, y = qD)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_point(aes(color = station)) + 
  scale_color_manual(values = station_colors) + 
  labs(x = "Salinity (PSU)", y = "Effective Number of ASVs") +
  stat_smooth(method = "lm") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```




