---
title: "Biodiversity across a salinity gradient"
author: "Marian Schmidt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../figures/05_Biodiversity")
```

# Goals 

Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the hill numbers as it relates to stations and salinity (in PSU).
4. Answer the scientific question at hand! 

## Inputs 

1. `phytree_preprocessed_physeq`

## Outputs 

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to visualize these changes in biodiversity
3. Statistical tests to actually show the changes and patterns in biodiversity.


# Set up the Environment 

## Load Packages & Colors 

```{r load-packages-colors}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               install = FALSE)

# Load in colors for plotting 
source("code/colors.R")
```

