---
title: "Microbial Composition Analysis Along A Salinity Gradient"
author: "Marian Schmidt"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goals

1. Load in scaled phyloseq object. 
2. Calculate relative abundances of taxonomic groups at various levels: 
    A. Phylum
    B. Family 
    C. Genus
    D. ASV 
3. Plot them, and narrow in on specific taxnomic group of interest

## Inputs 

1. scaled phyloseq object! 

## Outputs 

1. Beautiful visualizations of microbial taxa and how they vary across station/salinity.
2. Run some stats 

# Setup 

```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, 
               install = FALSE)

# load colors
source("code/colors.R")
```


# 1. Load in Scaled Phyloseq object 

```{r load-data}
load("data/06_Ordination/scaled_physeq.RData")

# Look at the data 
scaled_physeq

# Intuition check - scaled at 1,942
min(sample_sums(scaled_physeq))
range(sample_sums(scaled_physeq))
```

# Taxonomic Analysis! 

## Phylum 

Calculate the relative abundance of the phyla across all the samples. 

**NOTE!** The data MUST be normalized for this comparison. Here, we've scaled the read counts to 1,942.

```{r calc-phylum-df}

phylum_df <- 
  scaled_physeq %>%
  # Agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Phylum") %>%
  # Calculate the relative abundance! 
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # Create a dataframe from phyloseq object
  psmelt() %>%
  # Filter out Phyla < 1 % 
  dplyr::filter(Abundance > 0.01) %>%
   # fix the order of date
  mutate(date = fct_relevel(date, c("6/2/21", "6/15/21", "10/5/21")),
         station = fct_relevel(station, c("Copano West", "Copano East",
                                          "Mesquite Bay", "Aransas Bay",
                                          "Shipping Channel")))
```


## Stacked Bar plot 
```{r stacked-bar-phylum}
phylum_df %>%
  # Individual sample comparison of surface waters, whole fraction 
  # This will allow us to look at individual samples! 
  dplyr::filter(depth == 0.0) %>%
  dplyr::filter(fraction == "Whole") %>%
  ggplot(aes(x = station, y = Abundance, fill = Phylum)) + 
  facet_grid(Phylum~date) + 
  # add the stacked bar 
  geom_bar(stat = "identity", color = "black") + 
  # change the colors to be our selected colors 
  scale_fill_manual(values = phylum_colors)
```

# Actinomycetota 

```{r phylum-actinos}
phylum_df %>%
  dplyr::filter(Phylum == "Actinomycetota") %>%
  ggplot(aes(x = salinity_psu, y = Abundance, color = station, fill = station))+ 
  #geom_boxplot(alpha = 0.5, outlier.shape = NA)  + 
  geom_point() + 
  theme_bw() +
  scale_color_manual(values = station_colors) + 
  scale_fill_manual(values = station_colors) + 
  labs(title = "Actinomycetota Relative Abundance")

```



## Family 

## Genus 

## ASV 




